---
title: "Airline Data Analysis"
author: "Hamed Bastan-Hagh"
date: "6 July 2016"
output: html_document
---
A step-by-step account of data work done for this analysis of airline data.

Data were not available via URL, so the files were downloaded on 7 Jul 2016 at 
11:09. Changed them to xlsx format. Now read them into memory
```{r}
library(xlsx)
library(dplyr)
library(lubridate)
d <- read.xlsx("combinedenplane.xlsx", 1, startRow = 3, endRow = 199)
dsum <- d %>% 
    mutate(year = year(OBS_DATE)) %>% 
    select(year, OBS_DATE:ENPLANE_D11) %>% 
    group_by(year) %>% 
    summarise(total = sum(ENPLANE))
dsum <- tbl_df(dsum)

domd <- read.xlsx("domesticenplane.xlsx", 1, startRow = 3, endRow = 199)
domsum <- domd %>% 
    mutate(year = year(OBS_DATE)) %>% 
    select(year, OBS_DATE:ENPLANE_D_D11) %>% 
    group_by(year) %>% 
    summarise(domestic = sum(ENPLANE_D))
domsum <- tbl_df(domsum)

intd <- read.xlsx("intenplane.xlsx", 1, startRow = 3, endRow = 199)
intsum <- intd %>% 
    mutate(year = year(OBS_DATE)) %>% 
    select(year, OBS_DATE:ENPLANE_I_D11) %>% 
    group_by(year) %>% 
    summarise(intl = sum(ENPLANE_I))
intsum <- tbl_df(intsum)

airdata <- cbind(dsum, domsum$domestic, intsum$intl)
airdata <- tbl_df(airdata)
names(airdata) <- c("year", "total", "domestic", "intl")
airdata
```

The totals for 2008-14 tally with those from the World Bank and the International Civil Aviation Organization, but differ slightly from those in the MarketLine report. The differences 
are small enough that we can attribute this to methodology or adjustments that 
MarketLine may have made based on their own information. 

However the numbers for international and domestic are very different: the BTS 
numbers allocate much more of the total in each year to domestic flights. One 
clue as to why this might be comes from the instructions, which mention that "For the US and 
Canada, transborder passengers departing from either country are considered as 
part of the international segment". If the BTS allocate these to domestic that 
might explain the discrepancy.

I called and emailed the BTS but was unable to speak with any of the librarians at the National Transport Library. Therefore we cannot discern whether this explains the discrepancy. So to 'fix' the data I will use the 2015 numbers from the BTS for 
total passengers and allocate the proportions of domestic and international 
passengers to fit those in the MarketLine data, which averages at 79% domestic, 
21% international.

``` {r}
## Add columns with imputed values for domestic and international passengers
## to the airdata table
airdata2 <- airdata %>% 
    mutate(newdom = total * 0.79, newintl = total * 0.21) %>% 
    select(year, total, newdom, newintl)
airdata2
```

We can also use the BTS data for the first quarter of 2016 to get a sense of 
the year-on-year growth vs. 2015.

``` {r}
d2015q1 <- d %>% 
    mutate(year = year(OBS_DATE)) %>% 
    mutate(month = month(OBS_DATE)) %>% 
    select(year, month, OBS_DATE:ENPLANE) %>% 
    filter(year == 2015 & month <= 3) %>% 
    group_by(year) %>% 
    summarise(total = sum(ENPLANE))

d2016q1 <- d %>% 
    mutate(year = year(OBS_DATE)) %>% 
    mutate(month = month(OBS_DATE)) %>% 
    select(year, month, OBS_DATE:ENPLANE) %>% 
    filter(year == 2016) %>% 
    group_by(year) %>% 
    summarise(total = sum(ENPLANE))

d2 <- cbind(t(d2015q1), t(d2016q1))
diff <- d2[2, 2] / d2[2, 1]
diff
```

That means we've seen a c. 5% year-on-year increase from 2015 to 2016, and could 
expect about `r round(diff * airdata[16, 2] / 1000, 3)` million passengers in 
2016, of which c. `r round(diff * airdata[16, 2] * .79 / 1000, 3)` million would be 
domestic and `r round(diff * airdata[16, 2] * .21 / 1000, 3)` million would be
international.




